---
title: "Protein workflow"
author: "Petra Palenikova & Rick Scavetta"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Protein workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    fig.width=7, fig.height=4.5,
    collapse = TRUE,
    eval = TRUE,
    comment = "#>"
)
```

# Introduction

Complexome profiling or complexomics is a method used in biology to study 
migration profiles of proteins and protein complexes. First, sample is separated
into fractions, typically by blue native electrophoresis or gradient 
centrifugation and individual fractions are analysed by mass spectrometry. This
allows to identify proteins that co-migrate. Often people want to compare 
migration profiles and quantities of proteins between two biological samples. 
SILAC version of complexomics provides a useful tool for such studies. 

Here we present a package to analyse data produced by SILAC complexomics 
experiments. This package does not interpret raw mass spectrometry data. Protein
workflow of this package uses normalised protein data as an input. It allows to 
performs cluster analysis and contains tools for visualization of results. The
analysis is indented for samples that were SILAC labelled, therefore the input 
file should contain both 'heavy' and 'light' proteins.

### Method description

#### 0) Re-formatting of input file

Input file format was designed so it is easily readable by human. Downstream 
analysis requires a slightly different format so it is neccessary to perfomr 
this change before performaing the analysis.

#### 1) Hierarchical clustering

Clustering allows to identify similarity between migration profiles of proteins 
in an unbiased way. We can check co-migration of known protein complexes by 
simply filtering the data, however, clustering provides additional information
by allowing to identify unknown proteins that show similar migration profile as 
our proteins of interest.

This package contains functions to perform hierarchical clustering using Pearson
correlation (cantered or uncentered) as a distance measure and one of the three 
linkage methods (single, average or complete).

#### 2) Export files and visualizations

We provide several functions to export intermediate steps of the analysis.
Plotting functionality includes: 

* proteinPlot - line plot for a selected protein 
* groupHeatMap - heatmap for a selected group of proteins 
* oneGroupTwoLabelsCoMigration - scatter plot for a selected group of proteins
* twoGroupsWithinLabelCoMigration - scatter plot for 2 selected groups of
proteins
* makeBarPlotClusterSummary - bar plot showing number of proteins per cluster

# Example protein workflow

**Read in data, convert to correct format**

```{r}
library(ComPrAn)
inputFile <- system.file("extdata", "dataNormProts.txt", package = "ComPrAn")

forAnalysis <- protInportForAnalysis(data.table::fread(inputFile))
```


**Have a look at a selected protein (line plot)**
```{r}
protein <- "P52815"
max_frac <- 23
# example protein plot, quantitative comparison between labeled and unlabeled
# samples (default settings)
proteinPlot(forAnalysis[forAnalysis$scenario == "B",], protein, max_frac)
```

**Make a heatmap for a selected group of proteins**
```{r groupHeatMap, fig.width=7, fig.height=6.7}
groupDataFileName <- system.file("extdata","exampleGroup.txt",package="ComPrAn")
groupName <- 'group1'
groupData <- data.table::fread(groupDataFileName)
# example heatmap, quantitative comparison between labeled and unlabeled samples
# (default settings)
groupHeatMap(dataFrame = forAnalysis[forAnalysis$scenario == "B",],
                groupData, groupName)
```

**Co-migration plot of single protein group between label states**
```{r}
groupDataVector <- c("Q16540","P52815","P09001","Q13405","Q9H2W6")
groupName <- 'group1' 
max_frac <- 23 
# example co-migration plot, non-quantitative comparison of migration profile 
# of a sigle protein goup between labeled and unlabeled samples 
# (default settings)
oneGroupTwoLabelsCoMigration(forAnalysis, max_frac = max_frac, 
                                groupDataVector,groupName)

```

**Co-migration plot of two protein groups within label state**
```{r}
group1DataVector <- c("Q16540","P52815","P09001","Q13405","Q9H2W6")
group1Name <- 'group1' 
group2DataVector <- c("Q9NVS2","Q9NWU5","Q9NX20","Q9NYK5","Q9NZE8")
group2Name <- 'group2'
max_frac <- 23 
# example co-migration plot, non-quantitative comparison of migration profile 
# of two protein goups within label states (default settings)
twoGroupsWithinLabelCoMigration(dataFrame = forAnalysis, max_frac = max_frac, 
                                group1Data = group1DataVector, 
                                group1Name = group1Name,
                                group2Data = group2DataVector, 
                                group2Name = group2Name)
```

**Create components neccessary for clustering:**
(distance matrix for labeled and unlabeled samples,
protein table for both samples)
```{r}
clusteringDF <- clusterComp(forAnalysis,scenar = "A", PearsCor = "centered")
```

**Assign clusters to data frames**
```{r}
labTab_clust <- assignClusters(.listDf = clusteringDF,sample = "labeled",
                                    method = 'average', cutoff = 0.85)
unlabTab_clust <- assignClusters(.listDf = clusteringDF,sample = "unlabeled",
                                    method = 'average', cutoff = 0.85)
```

**Make bar plots** summarizing numbers of proteins per cluster for labeled
and unlabeled samples
```{r clusterBar, fig.width=4, fig.height=2.5}
makeBarPlotClusterSummary(labTab_clust, name = 'labeled')
makeBarPlotClusterSummary(unlabTab_clust, name = 'unlabeled')
```

**Create table containing proteins and their assigned clusters**
```{r}
tableForClusterExport <- exportClusterAssignments(labTab_clust,unlabTab_clust)
```

<center>
#### *End of file*
</center>